---
title: "Exploratory Factor Analysis (EFA)"
author: '*Adam Garber*'
subtitle: "A Course in `MplusAutomation`" 
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

### Load packages
```{r}
library(MplusAutomation)
library(tidyverse)
library(haven)
library(here)
library(gt) 
library(sjPlot)
```

### Load data example 

**Data source.** This tutorial utilizes the NCES public-use data called the Education Longitudinal Study of 2002 (Lauff & Ingels, 2014) This data can be found on the [$\color{blue}{\text{NCES website}}$](https://nces.ed.gov/surveys/els2002/avail_data.asp). Note that all examples used are for purposes of illustration only and are not intended to be interpreted substantively.

```{r}

els_data <-  read_spss("https://garberadamc.github.io/project-site/data/els_sub1_spss.sav")

```

### Prepare `data.frame` for analysis (select, reorder, & rename columns)
```{r}
schl_safe <- els_data %>% 
  select(
    'stu_tch' ="BYS20A", 'sc_spirt'="BYS20B", 
    'tch_good'="BYS20E", 'tch_intr'="BYS20F",
    'tch_prai'="BYS20G", 'stu_dwn' ="BYS20I",
    'not_safe'="BYS20J", 'disr_lrn'="BYS20K",
    'gangs'   ="BYS20M", 'rac_fght'="BYS20N",
    'sch_rule'="BYS21A", 'pun_same'="BYS21C",
    'strict'  ="BYS21D", 'pun_rule'="BYS21E",
    'female'  ="BYSEX",  'stu_race'="BYRACE", 
    'eng_natv'="BYSTLANG" 
  )
```

### View meta-data from labeled SPSS file
```{r, eval=FALSE}
# This meta-data or codebook can be downloaded as a PDF using the "print" option
sjPlot::view_df(schl_safe) 
```

### Look at variables for EFA example
```{r, echo=FALSE}
tribble(
~"Name", ~" Variable Description", 
#----------|-------------|,
"stu_tch"  , "Students get along well with teachers" ,
"sc_spirt" , "There is real school spirit" ,
"tch_good" , "The teaching is good" ,
"tch_intr" , "Teachers are interested in students" ,
"tch_prai" , "Teachers praise effort" ,
"stu_dwn"  , "In class often feels put down by students" ,
"not_safe" , "Does not feel safe at this school" ,
"disr_lrn" , "Disruptions get in way of learning" ,
"gangs"    , "There are gangs in school" ,
"rac_fght" , "Racial/ethnic groups often fight" ,
"sch_rule" , "Everyone knows what school rules are" ,
"pun_same" , "Punishment same no matter who you are" ,
"strict"   , "School rules are strictly enforced" ,
"pun_rule" , "Students know punishment for broken rules" ,
"female"   , "Student reported gender (male/female)" ,
"stu_race" , "Student reported race" ,
"eng_natv" , "Whether English is student's native language" ) %>% 
gt() %>% 
tab_header(
title = md("**Applied Example: School Safety**"), subtitle = md("&nbsp;")) %>%
tab_row_group(group = "", rows = 1:14) %>% 
tab_row_group(group = "Covariates", rows = 15:17) %>%
row_group_order(groups = c("","Covariates")) %>% 
tab_footnote(
  footnote = md("*Note*. All scale indicators have 4-point Likert response options ranging from *Strongly Agree (1)* to *Strongly Disagree (4)*."),
  locations = cells_title()) %>%
tab_options(column_labels.font.weight = "bold") 
```

 

### Reverse code indicators for factor interpretation

Expected factors based on theory and item similarity:

  - Factor 1: *School climate*, higher values indicate positive school climate
  - Factor 2: *safety*, higher values indicate safe school conditions
  - Factor 3: *clear rules*, higher values indicate clear communication of rules


```{r}
# select items to reverse code 
cols = c("stu_tch", "sc_spirt",                        # Factor 1: school climate 
         "tch_good", "tch_intr", "tch_prai",           # Factor 2: safety
         "sch_rule", "pun_same", "strict", "pun_rule") # Factor 3: clear rules
```



Use formula: `number of response categories` + `1` (e.g., `4 + 1` = `5`)
```{r}

 ### The number `5` in syntax below will change with applied context ###

schl_safe[ ,cols] <-  5 - schl_safe[ ,cols] 

```

 

### Check correlations to see if coding was correct (i.e., correlation is consistent within factor)

```{r}

library(corrplot)

f1_cor <- cor(schl_safe[1:7], use = "pairwise.complete.obs")

corrplot(f1_cor, 
         method = "circle",
         type = "upper",
         tl.col="black", 
         tl.srt=45)
```

***

### Create sub-folders for project organization:

1. create folder named `data`
2. create folder named `figures`
3. create folder named `efa_mplus`
4. create folder named `wls_efa`

***

### Prepare datasets

Prepare dataset for `mplusObject()` by removing SPSS labels 
```{r}

# write a CSV datafile (preferable format for reading into R, without labels)
write_csv(schl_safe, here("03-efa", "data", "els_efa_ready.csv"))

# read the unlabeled data back into R
efa_data <- read_csv(here("03-efa", "data", "els_efa_ready.csv"))

```

# --------------------------------------------------------------------------------------

### Estimate Exploratory Factor Analysis Model (EFA)

*Model 1: Default rotation*

# --------------------------------------------------------------------------------------

```{r, results="hide"}

efa_1  <- mplusObject(
  
  TITLE = "EFA", 
  
  VARIABLE = 
 "usevar =
  stu_tch sc_spirt tch_good tch_intr
  tch_prai stu_dwn not_safe disr_lrn 
  gangs rac_fght sch_rule pun_same strict pun_rule;",
  
  ANALYSIS = 
 "type = efa 1 5;   ! run efa of 1 through 5 factor models
  estimator = MLR;  ! using the ROBUST ML Estimator
  parallel=50;      ! run the parallel analysis for viewing elbow plot
  ",
 
  MODEL = "" ,
  
  PLOT = "type = plot3;",
 
  OUTPUT = "sampstat;",
  
  usevariables = colnames(efa_data), 
  rdata = efa_data)

efa_1_fit <- mplusModeler(efa_1, 
             dataout=here("03-efa","efa_mplus","efa_els.dat"),
             modelout=here("03-efa","efa_mplus","efa_els.inp"),
             check=TRUE, run = TRUE, hashfilename = FALSE)

```

 

### Create table summarizing model fit
```{r}

model_fit <- LatexSummaryTable(efa_1_fit,
  keepCols=c("Title","Parameters", "LL",
             "ChiSqM_Value","ChiSqM_DF", "ChiSqM_PValue",
             "RMSEA_Estimate","RMSEA_90CI_LB","RMSEA_90CI_UB",
             "CFI", "TLI","SRMR")) %>% 
  mutate(Title = c("1-Factor","2-Factor","3-Factor","4-Factor","5-Factor")) %>%
  mutate_at(vars(contains("RMSEA")), ~format(., nsmall = 3)) %>% 
  unite(CI, RMSEA_90CI_LB:RMSEA_90CI_UB, sep=", ", remove = TRUE) %>% 
  mutate(CI = paste0("(", CI, ")")) %>% 
  unite(RMSEA, RMSEA_Estimate:CI, sep=" ", remove = TRUE)
  
model_fit %>%
  gt() %>% 
  tab_header(
    title = md("**Table 1**"),
    subtitle = md("*Summary of Model Fit Indices*")) %>%
  cols_label(
    Title = "Model",
    Parameters = md("Par"),
    LL = md("*LL*"),
    ChiSqM_Value = md("Chi^2"),
    ChiSqM_PValue = md("*p-value*"),
    ChiSqM_DF = md("*df*"),
    RMSEA = "RMSEA (90% CI)" ) %>% 
  tab_options(column_labels.font.weight = "bold") %>%
  fmt(c(6), fns = function(x) ifelse(x<0.001,"<.001",
                                     scales::number(x, accuracy = 0.01)))
  
```


### Plot Parallel Analysis & Eigenvalues

Extract relevant data & prepare data.frame for plot
```{r}
x <- list(EFA=efa_1_fit$results$gh5$efa$eigenvalues, 
          Parallel=efa_1_fit$results$gh5$efa$parallel_average)

plot_data <- as_data_frame(x)
plot_data <- cbind(Factor = paste0(1:nrow(plot_data)), plot_data)

plot_data <- plot_data %>% 
  mutate(Factor = fct_inorder(Factor))
```

Pivot the dataframe to "long" format
```{r}
plot_data_long <- plot_data %>% 
  pivot_longer(EFA:Parallel,               # The columns I'm gathering together
               names_to = "Analysis",      # new column name for existing names
               values_to = "Eigenvalues")  # new column name to store values
```

Plot using ggplot
```{r}
plot_data_long %>% 
ggplot(aes(y=Eigenvalues,
           x=Factor,
           group=Analysis,
           color=Analysis)) +
  geom_point() +
  geom_line() + 
  theme_minimal() + 
  labs(title = "Figure 1: Parallel Eigenvalue Plot")
```

### save figure to the designated folder
```{r}
ggsave(here("03-efa","figures","eigenvalue_elbow_rplot.png"),
       dpi=300, height=5, width=7, units="in")
```


 

### Create table of EFA loading estimates
```{r}

loadings_stdyx <- efa_1_fit$results$parameters$efa$f3$loadings$estimates %>% 
   as.data.frame() %>% 
  rownames_to_column("Names") %>% 
  mutate(Names = str_to_lower(Names))

loadings_stdyx %>%
  gt() %>% 
  tab_header(
    title = md("**Table 2**"),
    subtitle = md("*Summary of Factor Loadings: 3-Factor EFA Model*")) %>%
  cols_label('1' = "F1", '2' = "F2", '3' = "F3") %>%
  tab_row_group(group = "Factor 1: School Climate", rows = 1:5) %>%
  tab_row_group(group = "Factor 2: Safety", rows = 6:10) %>%
  tab_row_group(group = "Factor 3: Clear Rules", rows = 11:14) %>%
  row_group_order(groups = c("Factor 1: School Climate","Factor 2: Safety","Factor 3: Clear Rules")) %>% 
  tab_style(style = list(cell_text(weight = "bold")),
            locations = cells_body(columns = "1",rows = 1:5)) %>% 
  tab_style(style = list(cell_text(weight = "bold")),
            locations = cells_body(columns = "2",rows = 6:10)) %>% 
  tab_style(style = list(cell_text(weight = "bold")),
            locations = cells_body(columns = "3",rows = 11:14)) %>%
  tab_options(column_labels.font.weight = "bold")

```

 

# --------------------------------------------------------------------------------------

### EFA with non-correlated factors

*Model 2: Varimax orthoginal rotation*

# --------------------------------------------------------------------------------------

```{r, results="hide"}

efa_2  <- mplusObject(
  TITLE = "EFA", 
  VARIABLE = 
 "usevar =
  stu_tch sc_spirt tch_good tch_intr
  tch_prai stu_dwn not_safe disr_lrn 
  gangs rac_fght sch_rule pun_same strict pun_rule;",
  
  ANALYSIS = 
 "type = efa 1 5;   
  estimator = MLR; 
  rotation = varimax;  !!! orthogonal (no factor correlations) !!!
  ",
 
  MODEL = "" ,
  
  PLOT = "type = plot3;",
 
  OUTPUT = "sampstat;",
  
  usevariables = colnames(efa_data), 
  rdata = efa_data)

efa_2_fit <- mplusModeler(efa_2, 
             dataout=here("03-efa","efa_mplus","m2_efa_els.dat"),
             modelout=here("03-efa","efa_mplus","m2_efa_els.inp"),
             check=TRUE, run = TRUE, hashfilename = FALSE)

```

Alternate syntax: use `update()` to alter the `mplusObject()` named `efa_1`

- `tilde` (~) will replace everything in that section of the input.
- `tilde-dot-plus` (~.+) will update the section by adding the specified code into that section

```{r, eval=FALSE}
efa_2  <- update(efa_1,
                 ANALYSIS = ~.+ "rotation = varimax;")

efa_2_fit <- mplusModeler(efa_2, 
             dataout=here("03-efa","efa_mplus","m2_efa_els.dat"),
             modelout=here("03-efa","efa_mplus","m2_efa_els.inp"),
             check=TRUE, run = TRUE, hashfilename = FALSE)

```

# --------------------------------------------------------------------------------------

## Applied example 2: EFA with Categorical indicators 

*Weighted Least Spuares Estimator* (WLS)

DATA SOURCE: This lab exercise utilizes a subset of the HSLS public-use dataset: High School Longitudinal Study of 2009 (Ingels et al., 2011) [$\color{blue}{\text{See website: nces.ed.gov}}$](https://nces.ed.gov/pubsearch/pubsinfo.asp?pubid=2018142)

```{r}

data_raw <- read_csv("https://garberadamc.github.io/project-site/data/hsls_fa_data_subset.csv")

```

## Reverse code for factor interpretation
```{r, eval=TRUE}

data_raw1 <- data_raw

cols = c("S1MPERS1", "S1MPERS2", "S1MUSELI", "S1MUSECL", "S1MUSEJO",
         "S1MTESTS", "S1MTEXTB", "S1MSKILL", "S1MASSEX", "S1MENJNG",
         "S1SPERS1", "S1SPERS2", "S1SUSELI", "S1SUSECL", "S1SUSEJO",
         "S1STESTS", "S1STEXTB", "S1SSKILL", "S1SASSEX", "S1SENJNG")

data_raw1[ ,cols] <-  5 - data_raw1[ ,cols]

```

### Prepare `data.frame` for analysis (select & rename columns)
```{r}
hsls_data <- data_raw1 %>% 
  select(
    'mth_pers'="S1MPERS1", 'mth_othr'="S1MPERS2", 'mth_life'="S1MUSELI",
    'mth_cllg'="S1MUSECL", 'mth_futr'="S1MUSEJO", 'mth_tsts'="S1MTESTS",
    'mth_text'="S1MTEXTB", 'mth_mstr'="S1MSKILL", 'mth_asgn'="S1MASSEX",
    'mth_enjy'="S1MENJNG", 'sci_pers'="S1SPERS1", 'sci_othr'="S1SPERS2",
    'sci_life'="S1SUSELI", 'sci_cllg'="S1SUSECL", 'sci_futr'="S1SUSEJO",
    'sci_tsts'="S1STESTS", 'sci_text'="S1STEXTB", 'sci_mstr'="S1SSKILL",
    'sci_asgn'="S1SASSEX", 'sci_enjy'="S1SENJNG")

```


### Look at variables for EFA example with categorical indicators
```{r, echo=FALSE}
tribble(
~"Name", ~" Variable Description", 
#----------|-------------|,
"mth_pers" , "9th grader sees himself/herself as a math person",
"mth_othr" , "Others see 9th grader as a math person",
"mth_life" , "9th grader thinks fall 2009 math course is useful for everyday life",
"mth_cllg" , "9th grader thinks fall 2009 math course will be useful for college",
"mth_futr" , "9th grader thinks fall 2009 math course is useful for future career",
"mth_tsts" , "9th grader confident can do excellent job on fall 2009 math tests",
"mth_text" , "9th grader certain can understand fall 2009 math textbook",
"mth_mstr" , "9th grader certain can master skills in fall 2009 math course",
"mth_asgn" , "9th grader confident can do excellent job on fall 2009 math assignments",
"mth_enjy" , "9th grader is enjoying fall 2009 math course very much",
"sci_pers" , "9th grader sees himself/herself as a science person",
"sci_othr" , "Others see 9th grader as a science person",
"sci_life" , "9th grader thinks fall 2009 science course is useful for everyday life",
"sci_cllg" , "9th grader thinks fall 2009 science course will be useful for college",
"sci_futr" , "9th grader thinks fall 2009 science course is useful for future career",
"sci_tsts" , "9th grader confident can do excellent job on fall 2009 science tests",
"sci_text" , "9th grader certain can understand fall 2009 science textbook",
"sci_mstr" , "9th grader certain can master skills in fall 2009 science course",
"sci_asgn" , "9th grader confident can do excellent job on fall 2009 science assignments",
"sci_enjy" , "9th grader is enjoying fall 2009 science course very much",
) %>% 
gt() %>% 
tab_header(
title = md("**Applied Example: Math & Science Utility**"), subtitle = md("&nbsp;")) %>%
tab_row_group(group = "Math Indicators", rows = 1:10) %>% 
tab_row_group(group = "Science Indicators", rows = 11:20) %>%
row_group_order(groups = c("Math Indicators","Science Indicators")) %>% 
tab_footnote(
  footnote = md("*Note*. All scale indicators have 4-point Likert response options ranging from *Strongly Agree (1)* to *Strongly Disagree (4)*."),
  locations = cells_title()) %>%
tab_options(column_labels.font.weight = "bold") 
```


# --------------------------------------------------------------------------------------

## Model 0 - Exploratory Factor Analysis (EFA) with WLS Estimator 

# --------------------------------------------------------------------------------------

```{r, results="hide"}

efa_wls  <- mplusObject(
  
  TITLE = 
    "EFA with Categorical Indicators - HSLS",
  
  VARIABLE = 
    "usevar = mth_pers-sci_enjy;
 
     categorical = mth_pers-sci_enjy;",
  
  ANALYSIS = 
    "type = efa 1 7;
     estimator=wlsmv;",
  
  MODEL = "" ,
  
  PLOT = "type = plot3;",
  
  OUTPUT = "sampstat;",
 
  usevariables = colnames(hsls_data), 
  rdata = hsls_data)

efa_wls_fit <- mplusModeler(efa_wls, 
                 dataout=here("03-efa", "wls_efa", "efa_sci_HSLS_wls.dat"),
                 modelout=here("03-efa", "wls_efa", "efa_sci_HSLS_wls.inp"),
                 check=TRUE, run = TRUE, hashfilename = FALSE)
```

 

### Create table summarizing model fit
```{r}

model_fit <- LatexSummaryTable(efa_wls_fit,
  keepCols=c("Title","Parameters", "LL",
             "ChiSqM_Value","ChiSqM_DF", "ChiSqM_PValue",
             "RMSEA_Estimate","RMSEA_90CI_LB","RMSEA_90CI_UB",
             "CFI", "TLI","SRMR")) %>% 
  mutate(Title = c("1-Factor","2-Factor","3-Factor",
                   "4-Factor","5-Factor","6-Factor","7-Factor")) %>%
  mutate_at(vars(contains("RMSEA")), ~format(., nsmall = 3)) %>% 
  unite(CI, RMSEA_90CI_LB:RMSEA_90CI_UB, sep=", ", remove = TRUE) %>% 
  mutate(CI = paste0("(", CI, ")")) %>% 
  unite(RMSEA, RMSEA_Estimate:CI, sep=" ", remove = TRUE)
  
model_fit %>%
  gt() %>% 
  tab_header(
    title = md("**Table 1**"),
    subtitle = md("*Summary of Model Fit Indices*")) %>%
  cols_label(
    Title = "Model",
    Parameters = md("Par"),
    #LL = md("*LL*"),
    ChiSqM_Value = md("Chi^2"),
    ChiSqM_PValue = md("*p-value*"),
    ChiSqM_DF = md("*df*"),
    RMSEA = "RMSEA (90% CI)" ) %>% 
  tab_options(column_labels.font.weight = "bold") %>%
  fmt(c(5), fns = function(x) ifelse(x<0.001,"<.001",
                                     scales::number(x, accuracy = 0.01)))
  
```

### Create table of EFA loading estimates
```{r}

loadings_stdyx <- efa_wls_fit$results$parameters$efa$f2$loadings$estimates %>% 
   as.data.frame() %>% 
  rownames_to_column("Names") %>% 
  mutate(Names = str_to_lower(Names))

loadings_stdyx %>%
  gt() %>% 
  tab_header(
    title = md("**Table 2**"),
    subtitle = md("*Summary of Factor Loadings: 2-Factor EFA Model*")) %>%
  cols_label('1' = "F1", '2' = "F2") %>%
  tab_row_group(group = "Factor 1: Math Indicators", rows = 1:10) %>%
  tab_row_group(group = "Factor 2: Science Indicators", rows = 11:20) %>%
  row_group_order(groups = c("Factor 1: Math Indicators",
                             "Factor 2: Science Indicators")) %>% 
  tab_style(style = list(cell_text(weight = "bold")),
            locations = cells_body(columns = "1",rows = 1:10)) %>% 
  tab_style(style = list(cell_text(weight = "bold")),
            locations = cells_body(columns = "2",rows = 11:20)) %>% 
  tab_options(column_labels.font.weight = "bold")

```

____________________________________

## References

Hallquist, M. N., & Wiley, J. F. (2018). MplusAutomation: An R Package for Facilitating Large-Scale Latent Variable Analyses in Mplus. Structural equation modeling: a multidisciplinary journal, 25(4), 621-638.

Muthén, L.K. and Muthén, B.O. (1998-2017).  Mplus User’s Guide.  Eighth Edition. Los Angeles, CA: Muthén & Muthén

R Core Team (2017). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL http://www.R-project.org/

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

# --------------------------------------------------------------------------------------

## Further resources & examples here:
  
$\color{blue}{\text{https://garberadamc.github.io/project-site/}}$

$\color{blue}{\text{https://www.adam-garber.com/}}$
  
# --------------------------------------------------------------------------------------

