---
title: "Confirmatory Factor Analysis (CFA)"
author: '*Adam Garber*'
subtitle: "A Course in `MplusAutomation`" 
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
library(here);library(glue)
```

# --------------------------------------------------------------------------------------

### Load packages
```{r}

library(MplusAutomation) # A conduit between R & Mplus
library(tidyverse);library(here);library(gt) # For tidiness & tables. 
library(semPlot);library(DiagrammeR) # For creating path diagrams.

```

# --------------------------------------------------------------------------------------

### Load data example 

**Data source.** This tutorial utilizes the NCES public-use data called the Education Longitudinal Study of 2002 (Lauff & Ingels, 2014) This data can be found on the [$\color{blue}{\text{NCES website}}$](https://nces.ed.gov/surveys/els2002/avail_data.asp). Note that all examples used are for purposes of illustration only and are not intended to be interpreted substantively.

```{r}

cfa_data <- read_csv("https://garberadamc.github.io/project-site/data/els_sub5_data.csv")

```

# --------------------------------------------------------------------------------------

### Prepare `data.frame` for analysis (select & reorder columns)
```{r}

school_trouble <- cfa_data %>% 
  select(stolen, t_hurt, p_fight, hit, damaged, bullied,  # factor 1 (indicators)
         safe, disrupt, gangs, rac_fght)                  # factor 2 (indicators)

```

### Look at variables for CFA example
```{r, echo=FALSE}
tribble(
~"Name", ~" Variable Description", 
#----------|-------------|,
"stolen"   , "Had something stolen at school"                   ,
"t_hurt"   , "Someone threatened to hurt 10th grader at school" ,
"p_fight"  , "Got into a physical fight at school"              ,
"hit"      , "Someone hit 10th grader"                          ,
"damaged"  , "Someone damaged belongings"                       ,
"bullied"  , "Someone bullied or picked on 10th grader"         ,
"safe"     , "Does not feel safe at this school"                ,
"disrupt"  , "Disruptions get in way of learning"               ,
"gangs"    , "There are gangs in school"                        ,
"rac_fght" , "Racial groups often fight"                        ) %>% 
gt() %>% 
tab_header(
title = md("**Applied Example: School Trouble**"), subtitle = md("&nbsp;")) %>%
tab_row_group(group = "Factor 1", rows = 1:6) %>%
tab_row_group(group = "Factor 2", rows = 7:10) %>%
row_group_order(groups = c("Factor 1","Factor 2")) %>% 
tab_options(column_labels.font.weight = "bold")
```



### Make a simple CFA path diagram using package {`DiagrammeR`} 
```{r}

grViz(" digraph cfa_model {

 graph [layout = dot, overlap = true]
 
 node [shape=box]
 stolen t_hurt p_fight hit damaged bullied safe disrupt gangs rac_fght;
 
 node [shape=circle, width = 0.9]
 factor1 factor2;
 
 edge []
 factor1 -> {stolen t_hurt p_fight hit damaged bullied}
 factor2 -> {safe disrupt gangs rac_fght}
 factor1 -> factor2 [dir = both, color = red]
 }")

```

# --------------------------------------------------------------------------------------

### Estimate CFA model 1 with default Unit Loading Identification (ULI)
```{r, results="hide"}

cfa_m1  <- mplusObject(
  
  TITLE = "CFA - ULI", 
  
  VARIABLE = 
    "usevar = stolen-rac_fght;", 
  
  ANALYSIS = 
    "estimator = mlr;",
  
  MODEL = 
    "factor1 by stolen t_hurt p_fight hit damaged bullied;
  
     factor2 by safe disrupt gangs rac_fght;" ,
  
  PLOT = 
    "type = plot3;",
  
  OUTPUT = 
    "sampstat standardized residual modindices (3.84);",
  
  usevariables = colnames(school_trouble), 
  rdata = school_trouble)

cfa_m1_fit <- mplusModeler(cfa_m1, 
              dataout = here("05-cfa","cfa_mplus","cfa_data.dat"),
              modelout = here("05-cfa","cfa_mplus","cfa_m1_uli.inp"),
              check=TRUE, run = TRUE, hashfilename = FALSE)
```



### Create a path diagram with `semPlot` package
```{r}

semPaths(cfa_m1_fit$results)

```



### Create table summarizing model fit
```{r}

model_fit <- LatexSummaryTable(cfa_m1_fit,
  keepCols=c("Title", "Parameters", "LL",
             "ChiSqM_Value","ChiSqM_DF", "ChiSqM_PValue",
             "RMSEA_Estimate","RMSEA_90CI_LB","RMSEA_90CI_UB",
             "CFI", "TLI","SRMR")) %>% 
  mutate(RMSEA_90CI_LB = str_pad(RMSEA_90CI_LB, width=6, side="left", pad="("),
         RMSEA_90CI_UB = str_pad(RMSEA_90CI_UB, width=6, side="right", pad=")")) %>% 
  unite(CI, RMSEA_90CI_LB:RMSEA_90CI_UB, sep=", ", remove = TRUE) %>% 
  unite(RMSEA, RMSEA_Estimate:CI, sep=" ", remove = TRUE)
  
model_fit %>%
  gt() %>% 
  tab_header(
    title = md("**Table 1**"),
    subtitle = md("*Summary of Model Fit Statistics*")) %>%
  cols_label(
    Title = "Model",
    Parameters = md("Par"),
    LL = md("*LL*"),
    ChiSqM_Value = md("Chi^2"),
    ChiSqM_PValue = md("*p-value*"),
    ChiSqM_DF = md("*df*"),
    RMSEA = "RMSEA (90% CI)" ) %>% 
  tab_options(column_labels.font.weight = "bold") %>%
  fmt(c(6), fns = function(x) ifelse(x<0.001,"<.001", scales::number(x, accuracy = 0.01)))
  
```



### Create table of CFA loading & correltation estimates 
```{r}

loadings_stdyx <- cfa_m1_fit$results$parameters$stdyx.standardized %>% 
  filter(grepl("FACTOR",paramHeader)) %>%
  select(param,est,se) %>% 
  mutate(param = str_replace(param, "FACTOR1", "F1 with F2"),
         param = str_to_title(param))

loadings_stdyx %>%
  gt() %>% 
  tab_header(
    title = md("**Table 2**"),
    subtitle = md("*Standardized Factor Loading and Factor Correlation Estimates*")) %>%
  cols_label(param = "Item", est = "Loading", se = "SE") %>%
  tab_row_group(group = "Factor 1", rows = 1:6) %>%
  tab_row_group(group = "Factor 2", rows = 7:10) %>%
  tab_row_group(group = "Factor Correlation", rows = 11) %>%
  row_group_order(groups = c("Factor 1","Factor 2","Factor Correlation")) %>% 
  tab_options(column_labels.font.weight = "bold")

```

# --------------------------------------------------------------------------------------

### Estimate CFA model (UVI; Unit Variance Identification)
```{r, results="hide"}

cfa_m2  <- mplusObject(
  
  TITLE = "CFA - Unit Variance Identification", 
  
  VARIABLE = 
    "usevar = stolen-rac_fght;", 
  
  ANALYSIS = 
    "estimator = mlr;",
  
  MODEL = 
    "!!! estimate first variable loading & fix variance at 1 !!!
     
     factor1 by stolen* t_hurt p_fight hit damaged bullied;
     factor1@1;
  
     factor2 BY safe* disrupt gangs rac_fght;
     factor2@1;",
  
  PLOT = 
    "type = plot3;",
  
  OUTPUT = 
    "sampstat standardized residual modindices (3.84);",
  
  usevariables = colnames(school_trouble), 
  rdata = school_trouble)

cfa_m2_fit <- mplusModeler(cfa_m1, 
              dataout = here("05-cfa","cfa_mplus","cfa_data.dat"),
              modelout = here("05-cfa","cfa_mplus","cfa_m2_uvi.inp"),
              check=TRUE, run = TRUE, hashfilename = FALSE)
```


### Create a path diagram including parameter estimates
```{r}
semPaths(cfa_m2_fit$results, "std", intercepts=FALSE,
         fixedStyle = c(1),
         fade = FALSE,
         color= list(lat = c("light blue"," light pink")),
)
```


## References

Hallquist, M. N., & Wiley, J. F. (2018). MplusAutomation: An R Package for Facilitating Large-Scale Latent Variable Analyses in Mplus. Structural equation modeling: a multidisciplinary journal, 25(4), 621-638.

Muthén, L.K. and Muthén, B.O. (1998-2017).  Mplus User’s Guide.  Eighth Edition. Los Angeles, CA: Muthén & Muthén

R Core Team (2017). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL http://www.R-project.org/

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

# --------------------------------------------------------------------------------------

## Further resources & examples here:
  
$\color{blue}{\text{https://garberadamc.github.io/project-site/}}$

$\color{blue}{\text{https://www.adam-garber.com/}}$
  
# --------------------------------------------------------------------------------------
