---
title: "Introduction to Mixture Models - Latent Class Analysis (LCA)"
author: '*Adam Garber*'
subtitle: "A Course in `MplusAutomation`" 
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
library(here);library(linguisticsdown)
```

# --------------------------------------------------------------------------------------


```{r, echo=FALSE, eval=TRUE, out.width = "60%", out.height= "60%", fig.pos="h"}

include_graphics2("https://garberadamc.github.io/project-site/figures/sem_frame_diagram.png")
```

*Figure*. Picture has been adapted from study by Múthen, 2006.

# --------------------------------------------------------------------------------------

## Preparation

# --------------------------------------------------------------------------------------

## Data source:

1. The first example utilizes data on undergraduate *Cheating* available from the `poLCA` package (Dayton, 1998): [$\color{blue}{\text{See documentation here}}$](https://cran.r-project.org/web/packages/poLCA/poLCA.pdf)

2. The second examples utilizes the public-use data, *The Longitudinal Survey of American Youth* (**LSAY**):  [$\color{blue}{\text{See documentation here}}$](https://www.lsay.org/)

3. The third examples utilizes data from a study evaluating the *Kindergarten Student Entrance Profile* (**KSEP**) (Quirk et al., 2011): [$\color{blue}{\text{See documentation here}}$](https://www.tandfonline.com/doi/full/10.1080/15377903.2010.540518?casa_token=hGkNrXYxjfAAAAAA:P4jjtIDXjHkU6mq8hvM9wsbQ-HgERqh0Z8cTfHIu4aeNUqThOtzKflpTwbWQpjFk9rT8At6-5Npw)

# --------------------------------------------------------------------------------------

Load packages
```{r, eval=TRUE}

library(tidyverse)
library(haven)
library(glue)
library(MplusAutomation)
library(here)
library(janitor)
library(gt)
library(reshape2)
library(cowplot) 
library(poLCA)

```

# --------------------------------------------------------------------------------------

## 1.1 Enumeration

Estimate *K*-class models with 1 through 6 classes.

# --------------------------------------------------------------------------------------

## Example 1 - Undergraduate Cheating behavior

"Dichotomous self-report responses by 319 undergraduates to four questions about cheating behavior" (poLCA, 2016).

# --------------------------------------------------------------------------------------

```{r, echo=FALSE, eval=TRUE, out.width = "50%", out.height= "50%", fig.pos="h"}

include_graphics2("https://garberadamc.github.io/project-site/figures/LCA_Diagram0_SEMLAB8.png")
```


```{r, echo=FALSE, eval=TRUE}

var_table0 <- tribble(
   ~"Name",      ~"Label",  ~"Values",                                   
#--------------|--------------------------------|-----|,
  "LieExam",   "lied to avoid taking an exam",  "0 = No, 1 = Yes",
  "LiePaper",  "lied to avoid handing a term paper in on time",  "0 = No, 1 = Yes",
  "Fraud", "purchased a term paper to hand in as their own or ...",  "0 = No, 1 = Yes",
  "CopyExam",   "copied answers during an exam from someone sitting near to them ",  "0 = No, 1 = Yes")

var_table0 %>% 
gt() %>% 
  tab_header(
    title = "LCA indicators",  # Add a title
    subtitle = md("&nbsp;")) %>%
  tab_options(
    table.width = pct(80)
  ) %>%
  tab_footnote(
    footnote = "Undergraduate Cheating Behavior",
    location = cells_title())

```

Prepare data 
```{r, eval=TRUE}

data(cheating)

cheating <- cheating %>% clean_names() 

df_cheat <-  cheating %>%                         
  dplyr::select(1:4) %>%                          
  dplyr::mutate_all(funs(.-1))                    

```


## Estimate a quick LCA model

**Specification details:**

- Within the `lapply()` function `1:4` indicates the number of *K*-class models to estimate.
- The measurement model indicator names for this example are listed after the `categorical = ...;` & `usevar = ...;` statements. 
- Note that in `Mplus` variables ordered as neighboring columns in the `data.frame` can be dplyr::selected by listing the first variable and last variable separated by a dash (e.g., `enjoy-adult`) . 

```{r, eval=FALSE}

lca_k1_4  <- lapply(1:4, function(k) {
  
  lca_enum  <- mplusObject(
      
    TITLE = glue("Class {k}"), 
  
    VARIABLE = glue(
    "categorical = lieexam-copyexam; 
     usevar = lieexam-copyexam;
     classes = c({k}); "),
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    starts = 200 100; 
    processors = 10;",
  
  OUTPUT = "tech11 tech14;",
  
  PLOT = 
    "type = plot3; 
    series = lieexam-copyexam(*);",
  
  usevariables = colnames(df_cheat),
  rdata = df_cheat)

lca_enum_fit <- mplusModeler(lca_enum, 
                            dataout=glue(here("12-intro-mixtures", "enum_cheat", "lca_cheat.dat")),
                            modelout=glue(here("12-intro-mixtures", "enum_cheat", "c{k}_lca_cheat.inp")) ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
})

```

# --------------------------------------------------------------------------------------

##  1.2 Generate Model Fit Summary Table 

- This syntax can be used to compare model fit from the series of LCA models generated during enumeration
- The code produces a table that is approximately in APA format. 

# --------------------------------------------------------------------------------------

Read in model fit statistics using `readModels()` and `mixtureSummaryTable()` functions
```{r}
output_cheat <- readModels(here("12-intro-mixtures", "enum_cheat"), quiet = TRUE)

enum_summary <- LatexSummaryTable(output_cheat,                                          
                keepCols=c("Title", "Parameters", "LL", "BIC", "aBIC",
                           "BLRT_PValue", "T11_VLMR_PValue","Observations"), 
                sortBy = "Title")
```

Calculate relevant fit indices for summary table
```{r}
allFit <- enum_summary %>% 
  mutate(aBIC = -2*LL+Parameters*log((Observations+2)/24)) %>% 
  mutate(CIAC = -2*LL+Parameters*(log(Observations)+1)) %>% 
  mutate(AWE = -2*LL+2*Parameters*(log(Observations)+1.5)) %>%
  mutate(SIC = -.5*BIC) %>% 
  mutate(expSIC = exp(SIC - max(SIC))) %>% 
  mutate(BF = exp(SIC-lead(SIC))) %>% 
  mutate(cmPk = expSIC/sum(expSIC)) %>% 
  dplyr::select(1:5,9:10,6:7,13,14) %>% 
  arrange(Parameters)
```


Generate the fit summary table
```{r}
allFit %>% 
  mutate(Title = str_remove(Title, " LCA Enumeration - Cheating Behavior Example")) %>% 
  gt() %>%
  tab_header(
    title = md("**Model Fit Summary Table**"), subtitle = md("&nbsp;")) %>% 
  cols_label(
    Title = "Classes",
    Parameters = md("Par"),
    LL = md("*LL*"),
    T11_VLMR_PValue = "VLMR",
    BLRT_PValue = "BLRT",
    BF = md("BF"),
    cmPk = md("*cmPk*")) %>%
  tab_footnote(
    footnote = md(
    "*Note.* Par = parameters; *LL* = log likelihood;
      BIC = bayesian information criterion;
      aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
      AWE = approximate weight of evidence criterion;
      BLRT = bootstrapped likelihood ratio test p-value;
      VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value;
      cmPk = approximate correct model probability."), 
    locations = cells_title()) %>% 
  tab_options(column_labels.font.weight = "bold") %>% 
  fmt_number(10,decimals = 2,
             drop_trailing_zeros=TRUE,
             suffixing = TRUE) %>% 
  fmt_number(c(3:9,11), 
             decimals = 0) %>% 
  fmt_missing(1:11,
              missing_text = "--") %>% 
  fmt(c(8:9,11),
    fns = function(x) 
    ifelse(x<0.001, "<.001",
           scales::number(x, accuracy = 0.01))) %>%
  fmt(10, fns = function(x) 
    ifelse(x>100, ">100",
           scales::number(x, accuracy = .1))) 

```


# --------------------------------------------------------------------------------------

##  1.3 Plot Information Criteria 

# --------------------------------------------------------------------------------------

```{r}
allFit %>% dplyr::select(2:7) %>% 
  rowid_to_column() %>% 
  pivot_longer(`BIC`:`AWE`, 
    names_to = "Index", 
    values_to = "ic_value") %>% 
  mutate(Index = factor(Index,
    levels = c("AWE","CIAC","BIC","aBIC"))) %>%
  ggplot(aes(x = rowid, y = ic_value,
    color = Index, shape = Index,
    group = Index, lty = Index)) + 
  geom_point(size = 2.0) + geom_line(size = .8) +
  scale_x_continuous(breaks = 1:6) +
  labs(x = "Number of Classes", y = "Information Criteria Value") +
  theme_cowplot() + theme(legend.title = element_blank(), legend.position = "top")
```

```{r}
ggsave(here("12-intro-mixtures", "figures", "Fig1_IC_plot_cheat.png"), dpi=300, height=5, width=7, units="in")
```

# --------------------------------------------------------------------------------------

##  1.4 Compare Conditional Item Probability Plots

# --------------------------------------------------------------------------------------

```{r, fig.width = 7, fig.height = 5}

model_results <- data.frame()
for (i in 1:length(output_cheat)) {
  temp <- data.frame(unclass(output_cheat[[i]]$parameters$unstandardized)) %>% 
    mutate(model = paste0(i, "-Class Model")) 
  model_results <- rbind(model_results, temp) }

pp_plots <- model_results %>% 
  filter(paramHeader == "Thresholds") %>% dplyr::select(est, model, LatentClass, param) %>% 
  mutate(prob = (1 / (1 + exp(est))), param = str_to_title(str_remove_all(param, "[$1]")))

ggplot(pp_plots,
       aes(x = param, y = prob, color = LatentClass, shape = LatentClass, group = LatentClass)) + 
  geom_point() + geom_line() + facet_wrap(~ model, ncol = 2) + labs(x= "", y = "Probability") +
  theme_minimal() + theme(legend.position = "none", axis.text.x = element_text(size = 6))
```

```{r}
ggsave(here("12-intro-mixtures", "figures","Fig2_compare_Kclass_Cheat.png"), dpi=300, height=4, width=6, units="in")
```

# --------------------------------------------------------------------------------------

##  1.5 Plot Final Model - Conditional Item Probability Plot 

# --------------------------------------------------------------------------------------

### This syntax creates a function called `plot_lca_function` that requires 7 arguments (inputs):

- `model_name`: name of Mplus model object (e.g., `model_step1`)
- `item_num`: the number of items in LCA measurement model (e.g., `5`)
- `class_num`: the number of classes (*k*) in LCA model (e.g., `3`)
- `item_labels`: the item labels for x-axis (e.g., `c("Enjoy","Useful","Logical","Job","Adult")`)
- `class_labels`: the class label names (e.g., `c("Adaptive Coping","Externalizing Behavior","No Coping")`)
- `class_legend_order` = change the order that class names are listed in the plot legend (e.g., `c(2,1,3)`)
- `plot_title`: include the title of the plot here (e.g., `"LCA Posterior Probability Plot"`)

Read in plot data from Mplus output file `c5_lca_enum.out`
```{r}

model_c4 <- readModels(here("12-intro-mixtures", "enum_cheat", "c4_lca_cheat.out"), quiet = TRUE)
                           
```

```{r}

plot_lca_function <- function(model_name,item_num,class_num,item_labels,
                              class_labels,class_legend_order,plot_title){

mplus_model <- as.data.frame(model_name$gh5$means_and_variances_data$estimated_probs$values)
plot_data <- mplus_model[seq(2, 2*item_num, 2),]

c_size <- as.data.frame(model_name$class_counts$modelEstimated$proportion)
colnames(c_size) <- paste0("cs")
c_size <- c_size %>% mutate(cs = round(cs*100, 2))
colnames(plot_data) <- paste0(class_labels, glue(" ({c_size[1:class_num,]}%)"))
plot_data <- plot_data %>% relocate(class_legend_order)

plot_data <- cbind(Var = paste0("U", 1:item_num), plot_data)
plot_data$Var <- factor(plot_data$Var,
               labels = item_labels)
plot_data$Var <- fct_inorder(plot_data$Var)

pd_long_data <- melt(plot_data, id.vars = "Var") 

# This syntax uses the data.frame created above to produce the plot with `ggplot()`

p <- pd_long_data %>%
  ggplot(aes(x = as.integer(Var), y = value,
  shape = variable, colour = variable, lty = variable)) +
  geom_point(size = 4) + geom_line() + 
  scale_x_continuous("", breaks = 1:item_num, labels = plot_data$Var) + 
  labs(title = plot_title, y = "Probability") +
  theme_cowplot() +
  theme(legend.title = element_blank(), 
        legend.position = "top")

p
return(p)
}

```


## Run the `plot_lca_function` by specifying each input (*Figure 1*)
```{r}

plot_lca_function(
  model_name = model_c4, 
  item_num = 4,
  class_num = 4,
  item_labels = c("Copy Exam", "Fraud", "Lie Exam", "Lie Paper"),
  class_labels = c("No Cheating","Copy/Fraud","Exam/Paper","Cheating"),
  class_legend_order = c(4,2,3,1),
  plot_title = "LCA Posterior Probability Plot - Cheating Behavior (K = 4)"
  )

```

```{r}
ggsave(here("12-intro-mixtures", "figures", "Fig3_C4_LCA_cheat.png"), dpi=300, height=5, width=7, units="in")
```


# --------------------------------------------------------------------------------------

## Example 2: Longitudinal Study of American Youth, **Science Attitudes** 

# --------------------------------------------------------------------------------------

```{r, echo=FALSE, eval=TRUE, out.width = "65%", out.height= "65%", fig.pos="h"}

include_graphics2("https://garberadamc.github.io/project-site/figures/LCA_Diagram_SEMLAB8.png")
```

Load data
```{r}

lsay_data <- read_csv("https://garberadamc.github.io/project-site/data/lca_lsay_sci.csv",
                      na = c("9999", "9999.00")) %>%               
  clean_names() %>%                                                                               
  dplyr::select(1:5, Enjoy = ab39m, Useful = ab39t,                                               
                     Logical = ab39u, Job = ab39w, Adult = ab39x)                                 


```

View LCA indicators 
```{r, echo=FALSE, eval=TRUE}

var_table <- tribble(
   ~"Name",      ~"Label",  ~"Values",                                   
#--------------|--------------------------------|-----|,
  "Enjoy",   "I enjoy science "                   ,  "0 = Disagree, 1 = Agree",
  "Useful",  "Science useful in everday problems ",  "0 = Disagree, 1 = Agree",
  "Logical", "Science helps logical thinkng "     ,  "0 = Disagree, 1 = Agree",
  "Job",     "Need science for a good job "       ,  "0 = Disagree, 1 = Agree",
  "Adult",   "Will use science often as an adult ",  "0 = Disagree, 1 = Agree")

gt(var_table) %>% 
  tab_header(
    title = "LCA Indicators",  # Add a title
    subtitle = md("&nbsp;")) %>% 
  tab_options(
    table.width = pct(80)
  ) %>%
  tab_footnote(
    footnote = "Longitudinal Study of American Youth",
    location = cells_title())


```

# --------------------------------------------------------------------------------------

##  2.1 Enumeration: 

Estimate *K*-class models with 1 through 6 classes.

# --------------------------------------------------------------------------------------

Run enumeration using `mplusObject` method
```{r, eval=FALSE}

lca_k1_6  <- lapply(1:6, function(k) {
  
  lca_enum  <- mplusObject(
      
    TITLE = glue("Class {k}"), 
  
    VARIABLE = glue(
    "categorical = Enjoy-Adult; 
     usevar = Enjoy-Adult;
     classes = c({k}); "),
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    starts = 200 100; 
    processors = 10;",
  
  OUTPUT = "sampstat residual tech11 tech14;",
  
  PLOT = 
    "type = plot3; 
    series = Enjoy-Adult(*);",
  
  usevariables = colnames(lsay_data),
  rdata = lsay_data)

lca_enum_fit <- mplusModeler(lca_enum, 
                            dataout=glue(here("12-intro-mixtures", "enum_lsay", "lca_lsay.dat")),
                            modelout=glue(here("12-intro-mixtures", "enum_lsay", "c{k}_lca.inp")) ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
})

```

# --------------------------------------------------------------------------------------

##  2.2 Generate Model Fit Summary Table 

# --------------------------------------------------------------------------------------

```{r}
output_lsay <- readModels(here("12-intro-mixtures", "enum_lsay"), quiet = TRUE)

enum_lsay <- LatexSummaryTable(output_lsay,                                 
                keepCols=c("Title", "Parameters", "LL", "BIC", "aBIC",
                           "BLRT_PValue", "T11_VLMR_PValue","Observations"),
                sortBy = "Title")
```

Calculate relevant fit indices for summary table
```{r}
allFit <- enum_lsay %>% 
  mutate(aBIC = -2*LL+Parameters*log((Observations+2)/24)) %>% 
  mutate(CIAC = -2*LL+Parameters*(log(Observations)+1)) %>% 
  mutate(AWE = -2*LL+2*Parameters*(log(Observations)+1.5)) %>%
  mutate(SIC = -.5*BIC) %>% 
  mutate(expSIC = exp(SIC - max(SIC))) %>% 
  mutate(BF = exp(SIC-lead(SIC))) %>% 
  mutate(cmPk = expSIC/sum(expSIC)) %>% 
  dplyr::select(1:5,9:10,6:7,13,14) %>% 
  arrange(Parameters)


```

Generate the fit summary table
```{r}
allFit %>% 
  mutate(Title = str_remove(Title, " LCA Enumeration - LSAY Example")) %>% 
  gt() %>%
  tab_header(
    title = md("**Model Fit Summary Table**"), subtitle = md("&nbsp;")) %>% 
  cols_label(
    Title = "Classes",
    Parameters = md("Par"),
    LL = md("*LL*"),
    T11_VLMR_PValue = "VLMR",
    BLRT_PValue = "BLRT",
    BF = md("BF"),
    cmPk = md("*cmPk*")) %>%
  tab_footnote(
    footnote = md(
    "*Note.* Par = parameters; *LL* = log likelihood;
      BIC = bayesian information criterion;
      aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
      AWE = approximate weight of evidence criterion;
      BLRT = bootstrapped likelihood ratio test p-value;
      VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value;
      cmPk = approximate correct model probability."), 
    locations = cells_title()) %>% 
  tab_options(column_labels.font.weight = "bold") %>% 
  fmt_number(10,decimals = 2,
             drop_trailing_zeros=TRUE,
             suffixing = TRUE) %>% 
  fmt_number(c(3:9,11), 
             decimals = 0) %>% 
  fmt_missing(1:11,
              missing_text = "--") %>% 
  fmt(c(8:9,11),
    fns = function(x) 
    ifelse(x<0.001, "<.001",
           scales::number(x, accuracy = 0.01))) %>%
  fmt(10, fns = function(x) 
    ifelse(x>100, ">100",
           scales::number(x, accuracy = .1))) 

```

# --------------------------------------------------------------------------------------

##  2.3 Plot Information Criteria 

# --------------------------------------------------------------------------------------

```{r}
allFit %>% dplyr::select(2:7) %>% 
  rowid_to_column() %>% 
  pivot_longer(`BIC`:`AWE`, 
    names_to = "Index", 
    values_to = "ic_value") %>% 
  mutate(Index = factor(Index,
    levels = c("AWE","CIAC","BIC","aBIC"))) %>%
  ggplot(aes(x = rowid, y = ic_value,
    color = Index, shape = Index,
    group = Index, lty = Index)) + 
  geom_point(size = 2.0) + geom_line(size = .8) +
  scale_x_continuous(breaks = 1:6) +
  labs(x = "Number of Classes", y = "Information Criteria Value") +
  theme_cowplot() + theme(legend.title = element_blank(), legend.position = "top")
```

```{r}
ggsave(here("12-intro-mixtures", "figures", "Fig1_IC_plot_LSAY.png"), dpi=300, height=5, width=7, units="in")
```

# --------------------------------------------------------------------------------------

##  2.4 Compare Conditional Item Probability Plots

# --------------------------------------------------------------------------------------

```{r, fig.width = 7, fig.height = 5}

model_results <- data.frame()
for (i in 1:length(output_lsay)) {
  temp <- data.frame(unclass(output_lsay[[i]]$parameters$unstandardized)) %>% 
    mutate(model = paste0(i, "-Class Model")) 
  model_results <- rbind(model_results, temp) }

pp_plots <- model_results %>% 
  filter(paramHeader == "Thresholds") %>% dplyr::select(est, model, LatentClass, param) %>% 
  mutate(prob = (1 / (1 + exp(est))), param = str_to_title(str_remove_all(param, "[$1]")))

ggplot(pp_plots,
       aes(x = param, y = prob, color = LatentClass, shape = LatentClass, group = LatentClass)) + 
  geom_point() + geom_line() + facet_wrap(~ model, ncol = 2) + labs(x= "", y = "Probability") +
  theme_minimal() + theme(legend.position = "none", axis.text.x = element_text(size = 6))
```

```{r}
ggsave(here("12-intro-mixtures", "figures","Fig2_compare_Kclass_LSAY.png"), dpi=300, height=4, width=6, units="in")
```

# --------------------------------------------------------------------------------------

##  2.5 Plot Final Model - Conditional Item Probability Plot 

# --------------------------------------------------------------------------------------

```{r}

model_c4 <- readModels(here("12-intro-mixtures", "enum_lsay", "c4_lca.out"), quiet = TRUE)
                           
```

## Run the `plot_lca_function` by specifying each input (*Figure 1*)
```{r}

plot_lca_function(
  model_name = model_c4, 
  item_num = 5,
  class_num = 4,
  item_labels = c("Enjoy", "Useful", "Logical", "Job", "Adult"),
  class_labels = c("Class-1","Class-2","Class-3","Class-4"),
  class_legend_order = c(1,2,3,4),
  plot_title = "LCA Posterior Probability Plot"
  )

```

# --------------------------------------------------------------------------------------

## Example 3 - Kindergarten Student Entrance Profile (**KSEP**; Quirk et al., 2011) 

# --------------------------------------------------------------------------------------


```{r}

ksep <- read_csv("https://garberadamc.github.io/project-site/data/ksep_sub_18.csv")

```


```{r, eval=TRUE, echo=FALSE}

var_table2 <- tribble(
   ~"Name",      ~"Label",  ~"Values",                                   
#--------------|--------------------------------|-----|,
  "seek_hlp" ,  "Seeks adult help when appropriate"  ,  "0 = Not Mastered, 1 = Mastered",
  "cooperat" ,  "Engages in cooperative play activities with peers"  ,  "0 = Not Mastered, 1 = Mastered",
  "imp_cntr" ,  "Exhibits impulse control and self-regulation"  ,  "0 = Not Mastered, 1 = Mastered",
  "repeats"  ,  "Stays with or repeats a task"  ,  "0 = Not Mastered, 1 = Mastered",
  "separate" ,  "Separates appropriately from caregiver most days"  ,  "0 = Not Mastered, 1 = Mastered",
  "new_activ",  "Is enthusiastic and curious in approaching new activities"  ,  "0 = Not Mastered, 1 = Mastered",
  "folw_rul" ,  "Follows rules when participating in routine activities"  ,  "0 = Not Mastered, 1 = Mastered",
  "name"     ,  "Recognizes own name"  ,  "0 = Not Mastered, 1 = Mastered",
  "writes"   ,  "Writes own name"  ,  "0 = Not Mastered, 1 = Mastered",
  "express"  ,  "Demonstrates expressive abilities"  ,  "0 = Not Mastered, 1 = Mastered",
  "quantity" ,  "Understands that numbers represent quantity"  ,  "0 = Not Mastered, 1 = Mastered",
  "colors"   ,  "Recognizes Colors"  ,  "0 = Not Mastered, 1 = Mastered",
  "shapes"   ,  "Recognizes primary shapes"  ,  "0 = Not Mastered, 1 = Mastered")

gt(var_table2) %>% 
  tab_header(
    title = "LCA Indicators - KSEP Example",
    subtitle = md("&nbsp;")) %>%
  tab_options(
    table.width = pct(80)
  ) %>%
  tab_footnote(
    footnote = "Kindergarten Student Entrance Profile",
    location = cells_title()) 
```

# --------------------------------------------------------------------------------------

##  3.1 Enumeration: 

Estimate *K*-class models with 1 through 6 classes.

# --------------------------------------------------------------------------------------

```{r, eval=FALSE}

lca_k1_6  <- lapply(1:6, function(k) {
  
  lca_enum  <- mplusObject(
      
    TITLE = glue("Class {k}"), 
  
    VARIABLE = glue(
    "categorical = seek_hlp-shapes; 
     usevar = seek_hlp-shapes;
     classes = c({k}); "),
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    !stseed = 5212020;
    starts = 200 100; 
    processors = 10;",
  
  OUTPUT = "sampstat residual tech11 tech14;",
  
  PLOT = 
    "type = plot3; 
    series = seek_hlp-shapes(*);",
  
  usevariables = colnames(ksep),
  rdata = ksep)

lca_enum_fit <- mplusModeler(lca_enum, 
                            dataout=glue(here("12-intro-mixtures", "enum_ksep", "lca_ksep.dat")),
                            modelout=glue(here("12-intro-mixtures", "enum_ksep", "c{k}_lca_ksep.inp")) ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
})

```

# --------------------------------------------------------------------------------------

##  3.5 Plot Final Model - Conditional Item Probability Plot 

# --------------------------------------------------------------------------------------

```{r}

model_c4 <- readModels(here("12-intro-mixtures", "enum_ksep", "c4_lca_ksep.out"), quiet = TRUE)
                           
```

## Run the `plot_lca_function` by specifying each input (*Figure 1*)
```{r}

plot_lca_function(
  model_name = model_c4, 
  item_num = 13,
  class_num = 4,
  item_labels = c("Seeks help", "Cooperative", "Impulse control", "Repeats",
                  "Separates", "New activities", "Follows rules", "Name",
                  "Writes", "Expressive", "Quantity", "Colors", "Shapes"),
  class_labels = c("C2","C1","C3","C4"),
  class_legend_order = c(2,1,3,4),
  plot_title = "LCA Posterior Probability Plot"
  )

```


```{r, eval=FALSE}
ggsave(here("12-intro-mixtures", "figures","C4_KSEP_LCA_plot.png"),    
       dpi=300, height=4, width=6, units="in")    
```

## References

Drew A. Linzer, Jeffrey B. Lewis (2011). poLCA: An R Package for Polytomous Variable Latent Class Analysis. Journal of Statistical Software, 42(10), 1-29. URL http://www.jstatsoft.org/v42/i10/.

Hallquist, M. N., & Wiley, J. F. (2018). MplusAutomation: An R Package for Facilitating Large-Scale Latent Variable Analyses in Mplus. Structural equation modeling: a multidisciplinary journal, 25(4), 621-638.

Miller, J. D., Hoffer, T., Suchner, R., Brown, K., & Nelson, C. (1992). LSAY codebook. Northern Illinois University.

Muthén, B. O., Muthén, L. K., & Asparouhov, T. (2017). Regression and mediation analysis using Mplus. Los Angeles, CA: Muthén & Muthén.

Muthén, L.K. and Muthén, B.O. (1998-2017).  Mplus User’s Guide.  Eighth Edition. Los Angeles, CA: Muthén & Muthén

Quirk, M., Furlong, M., Lilles, E., Felix, E., & Chin, J. (2011). Preliminary development of a kindergarten school readiness assessment for Latino students. Journal of Applied School Psychology, 27(1), 77-102.

R Core Team (2017). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL http://www.R-project.org/

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

# --------------------------------------------------------------------------------------

## Further resources & examples here:
  
$\color{blue}{\text{https://garberadamc.github.io/project-site/}}$

$\color{blue}{\text{https://www.adam-garber.com/}}$
  
# --------------------------------------------------------------------------------------